manuscript

### 2023-08-08 manuscript.rmd
Update manuscript.rmd
added to intro

@@

The ocean is naturally a noisy place, but recent increases in human activities are increasing the ocean sound levels (@tyack2008 ). Frequencies of these activities often overlap in frequency range, space, and time. Climate change is also expected to alter the temperature and salinity of the ocean, causing an increase in global sound speed @affatati2022 .

Ships account for over 80% of the global trade, and an estimated 2.1% annual increase in shipping in expected (@unitedstatesarmycorpsofengineersusace2018) . Shipping accounts for the most prominent anthropogenic noise in the ocean, so sound levels are expected to increase. \[Discuss post-covid elevated levels. \] The global ambient noise in the ocean is estimated to have increased by 12 decibels in the last few decades @hildebrand2009.

Within their Marine Strategy Framework Directive, the European Union has recognized noise as a form of pollution @directiv2008.

**\[Generally, how are cetaceans/fishes/invertebrates affected?\]**

Sound is a critical form of communication for many species in the ocean.

**\[Sound sources\]**\
\@pijanowski2011 (+ mapping soundscapes). Soundscapes are made up of three broad categories of sound sources:\
- biophonies, which come from marine mammals, fishes, and invertebrates;\
- geophonies, which consist of natural sounds like rain, wind, earthquakes;\
- anthropophonies, which consist of human-made sounds from vessels (@miksis-olds, @duarte2021) . These can be incidental or deliberate @chou2021, @chahouri2022 . Incidental = shipping, offshore development, drilling. Deliberate = active sonar pingers/echosounders, and seismic surveys.

**\[Chronic (long-term)/residual/ambient/transient/continuous/impulsive/acute sound\]** @mckenna2021 and @richardson2013\
Transient sounds have a short duration, while continuous sounds keep going. @richardson2013. How are transient and continuous sounds known to affect marine mammals differently? Close approach of a vessel, for example, appears to disrupt beaked whale foraging, as opposed to some medium levels of chronic ship noise ( @aguilarsoto2006 ). It is unknown if short-term effects translate into long-term effects

Lower frequencies travel further, and higher frequencies tend to attenuate (@richardson2013 ), so the focus should be on local effects for that (@hildebrand2009 ). Ambient noise in shallow water has more variability than deeper water (@richardson2013)

"Ambient noise off the coast of California has increased many-fold over the past several decades... (@moore2013 + other sources w/in that). It's likely this noise pollution will interfere with soniferous animals, but it's unclear how, where, and what species will be affected. \[add language about sensitivity and vulnerability @duarte2021?? i think).

@@

**\[Introduce beaked and sperm whales: protected, visually cryptic, animals that are sensitive to noise and use sound extensively to communicate\]**\
[Beaked whales]{.underline} (*Ziphiidae*) routinely dive for hours at a time to depths greater than 1000 ft. They can dive for over 3 hours, followed by only a few minutes on the surface (same for BW and SW?) ( see Watkins et al., 1993). This diving behavior, coupled with a low-profile body makes them nearly invisible on the surface, so very little is known about them. This may be an anti-predation adaptation to avoid killer whales (cite) as well as a synchronized foraging dives following a silent descent (@aguilardesoto2020 ). Because they must make sound to locate prey, they do so in the refuge of deep depths where killer whales do not go (less than 20 m @aguilardesoto2020 and in frequencies they do not hear (cite)). Why do they have both adaptations? It is thought that this anti-predator response is exactly what makes them sensitive to human noise.

@@

Species studied:\
- ZC Cuvier's beaked whale *Ziphius cavirostris (IUCN status: least concern)\
-* BW43, possibly Perrin's beaked whale *M*. *perrini (IUCN status: endangered)*\
- BW37V, possibly Hubbs' beaked whale *M. carlhubbsi (IUCN status: data deficient)\
-* BB, --???\
- BWC, possibly ginkgo-toothed beaked whale *M. ginkgodens (IUCN status: data deficient)\
-* PM, sperm whale *(IUCN status: vulnerable)*

@@

Marine Strategy Framework Directive by European Union establishes a directive about environmental regulation on marine noise, specifically requiring member states to adopt strategies in compliance with threshold values for impulsive and continuous noise. @directiv2008, @borsanij.f.2023

@@

Both of these outcomes are of high importance to managers and agencies especially NOAA and BOEM (see 2023-2024 report). Within their Ocean Noise Strategy Roadmap, NOAA is specifically seeking 1) a quantification of the spatial and temporal variability of ambient noise conditions, and 2) understanding of how anthropogenic soudn sources contribute to the soundscape. Co-occurrence of buoys and vessels will provide fine-scale snapshots of sound levels surrounding vessels and concurrent whale foraging activity. With AIS, we can calculate distance from buoy to vessel and sound levels in those moments. And 3) NOAA is seeking more information on vocally active species' distributions. This is part of a bill introduced by Congress to "assess underwater sound in high-priority" environments (https://www.congress.gov/bill/117th-congress/house-bill/6987).

@@

\
The California Current Ecosystem extends along the US' Economic Exclusion Zone along the west coast. It is characterized by... It's a highly productive and complex ecosystem that contains five current and one proposed national marine sanctuaries, supports several wild fisheries, and is home to many protected species. Because this area has high economic value and natural ecological activity, it's a well-managed and well-researched area.

@@

**\[Post-processing details, data cleaning\]**\
Soundscape metrics were rounded to the nearest 20 minutes (:00, :20, :40) and converted to UTC. These include third-octave level bands and broadband (20-24,000 Hz) metrics. ~~One way to measure sound is with a metric called a third octave level, which splits the sound spectrum into many bands.~~ The band at 125 Hertz is often associated with shipping vessel noise and the 2-5 kHz range with mid-frequency naval sonar ( @garrett2016 ). Beaked whales are known to be sensitive to sonar in this latter range (cite). Several TOL levels were chosen as proxies to vessel or sonar frequencies (from TOL 63 to 20,000).

@@

| Covariate                     | Unit | Source                                                          | Justification                                                                                                         |
|-------------------------------|------|-----------------------------------------------------------------|-----------------------------------------------------------------------------------------------------------------------|
|                               |      | \[can use A, B, C to avoid repetition, cite in caption\]        |                                                                                                                       |
| Distance to continental slope |      | General Bathymetric Chart of the Oceans (GEBCO) @weatherall2015 |                                                                                                                       |
| Bathymetric slope             |      | GEBCO tif                                                       |                                                                                                                       |
|                               |      |                                                                 |                                                                                                                       |
| Wind curl stress              |      | ERDDAP                                                          | "Positive wind stress curl lifts the thermocline, which results in enhanced primary production" (Manocci et al 2015). |
|                               |      |                                                                 |                                                                                                                       |
|                               |      |                                                                 |                                                                                                                       |
|                               |      |                                                                 |                                                                                                                       |
|                               |      |                                                                 |                                                                                                                       |
|                               |      |                                                                 |                                                                                                                       |
|                               |      |                                                                 |                                                                                                                       |
|                               |      |                                                                 |                                                                                                                       |
|                               |      |                                                                 |                                                                                                                       |
|                               |      |                                                                 |                                                                                                                       |

@@

>>>>

Sound is a critical form of communication for many species in the ocean. Since sound travels over four times faster in water than in air and can carry information over far distances @urick1983 , the use of sound has evolved as a critical component of communication, navigation, and foraging for many invertebrates, fishes and cetaceans @duarte2021 . Under proper condition, sounds have the ability to travel great distances across ocean basins @miksis-olds.

The total collection of sounds in an environment is called a soundscape, and these are grouped into three broad categories: biophonies, sound produced by cetaceans, fishes, and invertebrates; geophonies, which consist of natural sounds like rain, wind, and earthquakes; anthropophonies, which consist of human-made sounds from shipping vessels, sonar, and survey activities (@miksis-olds, @duarte2021). Ocean sound levels exhibit high spatio-temporal variation based on traits such as properties of water, organism diversity, human activities, and bathymetry of a region @andré2011. Within soundscape ecology, ocean soundscapes have unique enough sound signatures to be characterized, measured, and monitored as indicators of ecosystem health, species diversity @radford2010 @haver2019.

**\[what do they have to lose in a noisy environment?\]**

In an exceptionally noisy environment, organisms can be negatively impacted. Invertebrates and fishes may have stunted development and elevated stress responses (cite), while cetaceans' behavior, calling patterns, physiology and stress levels, ability to locate prey or conspecifics are affected. Ecosystem-level effects? Each species is affected differently, some more sensitive to certain types of noise. While terrestrial ecosystem-level effects have been documented, (Buxton 2017, Shannon 2016, Francis 2012), those within aquatic ecosystems are less studied @kunc2016.

@@

Ocean noise pollution is among the anthropogenic changes that have recently grown to concerning levels @tyack2008, @hildebrand. The global ambient noise in the ocean is estimated to have increased by 12 decibels in the last few decades and has a close correlation to global economic trends @frisk2012. with commercial shipping accounting for most of the low-frequency anthropogenic noise in the ocean @hildebrand2009. Ships have been increasing in terms of gross tonnage and fleet size @hildebrand2009 (and age?), and now account for over 80% of the global trade, with an estimated 1.4% expected annual increase in shipping through 2027 @unctad.

These noise-generating activities are categorized as either incidental or deliberate @chou2021, @chahouri2022. Deliberate activities such as active sonar pingers, resource extraction from the seabed, seismic surveys, and echosounders are intentionally produce sound to measure an area and can utilize both high and low frequencies. Incidental activities like dredging, offshore development, and shipping produce sound as a by-product @hawkins2017. Anthropogenic noise can further be characterized by the frequency and duration of the sound. For example, shipping has a low-frequency (\<200 Hz), but continuous sound that is widely distributed across the world's oceans @hildebrand2009. Because this type of noise can travel great distances (add number), they can have effects further away from the source. Sonar for military surveillance or research such as seafloor mapping occupies the mid-frequency bands, but attenuates quicker and occupies a smaller acoustic space than shipping (shorter distance). Close approach of a vessel, for example, appears to disrupt beaked whale foraging, as opposed to some medium levels of chronic ship noise ( @aguilarsoto2006 ).

It remains unknown if chronic effects from continuous sound or acute effects from impulsive sounds can alter the normal behavior or physiology of cetaceans enough to affect reproduction and population levels.

Lower frequencies travel further, and higher frequencies tend to attenuate (@richardson2013 ), so the focus should be on local effects for that (@hildebrand2009 ). Ambient noise in shallow water has more variability than deeper water (@richardson2013). Climate change is also expected to alter the temperature, salinity, and pH of the ocean, causing an additional increase in global sound speed due to changing absorption qualities @affatati2022 @kunc2016 .

**\[This is an issue, but the extent of damage remains unclear, so it's difficult to produce legislation to that effect\]**

As soniferous animals, cetaceans have varying levels of sensitivity and vulnerability to certain types of noise. Mysticetes are more affected by low-frequency noise produced by large shipping vessels. Odontocetes produce and hear mid- to high-frequency sounds, so human activities that produce sounds at these frequencies interfere with these signals. While many studies document negative responses of cetaceans to noise pollution with general frequency bands (cite), the circumstances surrounding these acoustic disturbances (chronic or impulsive, low, mid, or high frequency, shallow or deep water, overlap with anthropogenic activity) and their long-term effects are unclear.

@@

**\[Introduce beaked and sperm whales: protected, visually cryptic, animals that are sensitive to noise and use sound extensively to communicate\]**\
[Beaked whales]{.underline} (*Ziphiidae*) are among the least studied and most sensitive animals to anthropogenic noise. They routinely dive for hours at a time to depths greater than 1000 ft. They can dive for over 3 hours, followed by only a few minutes on the surface (same for BW and SW?) ( see Watkins et al., 1993). This diving behavior, coupled with a low-profile body makes them nearly invisible on the surface, so very little is known about them. This may be an anti-predation adaptation to avoid killer whales (cite) as well as a synchronized foraging dives following a silent descent (@aguilardesoto2020 ). Because they must make sound to locate prey, they do so in the refuge of deep depths where killer whales do not go (less than 20 m @aguilardesoto2020 and in frequencies they do not hear (cite)). Why do they have both adaptations? It is thought that this anti-predator response is exactly what makes them sensitive to human noise.

@@

Species studied:\
- ZC Cuvier's beaked whale *Ziphius cavirostris (IUCN status: least concern)\
-* BW43, possibly Perrin's beaked whale *M*. *perrini (IUCN status: endangered)* confirmed??\
- BW37V, possibly Hubbs' beaked whale *M. carlhubbsi (IUCN status: data deficient)\
-* BB, --???\
- BWC, possibly ginkgo-toothed beaked whale *M. ginkgodens (IUCN status: data deficient)\
-* PM, sperm whale *(IUCN status: vulnerable)*

@@

Marine Strategy Framework Directive by European Union establishes a directive about environmental regulation on marine noise, specifically requiring member states to adopt strategies in compliance with threshold values for impulsive and continuous noise. @directiv2008, @borsanij.f.2023 Within their Marine Strategy Framework Directive, the European Union is the only entity to date that has recognized noise as a form of pollution and restricts noise from reaching levels that may cause negative effects to the environment @directiv2008. [What about UK Marine Strategy?] See @chou2021 for larger list of efforts.

@@

Both of these outcomes are of high importance to managers and agencies especially NOAA and BOEM (see 2023-2024 report). Within their Ocean Noise Strategy Roadmap, NOAA is specifically seeking 1) a quantification of the spatial and temporal variability of ambient noise conditions, and 2) understanding of how anthropogenic sound sources contribute to the soundscape. Co-occurrence of buoys and vessels will provide fine-scale snapshots of sound levels surrounding vessels and concurrent whale foraging activity. With AIS, we can calculate distance from buoy to vessel and sound levels in those moments. And 3) NOAA is seeking more information on vocally active species' distributions. This is part of a bill introduced by Congress to "assess underwater sound in high-priority" environments (https://www.congress.gov/bill/117th-congress/house-bill/6987).

@@

\
The California Current Ecosystem extends along the US' Economic Exclusion Zone along the west coast. It is characterized by... It's a highly productive and complex ecosystem that contains five current and one proposed national marine sanctuaries, supports several wild fisheries, and is home to many protected species. Because this area has high economic value and natural ecological activity, it's a well-managed and well-researched area. "Ambient noise off the coast of California has increased many-fold over the past several decades..." (@moore2013 + other sources w/in that).

@@

**\[Post-processing details, data cleaning\]**\
Soundscape metrics were rounded to the nearest 20 minutes (:00, :20, :40) and converted to UTC. These include third-octave level bands and broadband (20-24,000 Hz) metrics. ~~One way to measure sound is with a metric called a third octave level, which splits the sound spectrum into many bands.~~ The band at 125 Hertz is often associated with shipping vessel noise and the 2-5 kHz range with mid-frequency naval sonar ( @garrett2016 ). Beaked whales are known to be sensitive to sonar in this latter range (cite). Several TOL levels were chosen as proxies to vessel or sonar frequencies, ranging from TOL 63 to 20,000.

@@

| Covariate                     | Unit | Source                                                          | Justification                                                                                                         |
|-----------------|-----------------|-----------------|----------------------|
|                               |      | \[can use A, B, C to avoid repetition, cite in caption\]        |                                                                                                                       |
| Distance to continental slope |      | General Bathymetric Chart of the Oceans (GEBCO) @weatherall2015 |                                                                                                                       |
| Bathymetric slope             |      | GEBCO tif                                                       |                                                                                                                       |
|                               |      |                                                                 |                                                                                                                       |
| Wind curl stress              |      | ERDDAP                                                          | "Positive wind stress curl lifts the thermocline, which results in enhanced primary production" (Manocci et al 2015). |
|                               |      |                                                                 |                                                                                                                       |
|                               |      |                                                                 |                                                                                                                       |
|                               |      |                                                                 |                                                                                                                       |
|                               |      |                                                                 |                                                                                                                       |
|                               |      |                                                                 |                                                                                                                       |
|                               |      |                                                                 |                                                                                                                       |
|                               |      |                                                                 |                                                                                                                       |
|                               |      |                                                                 |                                                                                                                       |
|                               |      |                                                                 |                                                                                                                       |

@@

AIS was downloaded, processed (converted to UTC), and matched to each drifter using Taiki's PAMscapes package and functions.
